Global sensitivity analysis of model uncertainty in aeroelastic wind turbine models

### 2020-06-02 Papers/Torque_Paper/revision/content.tex
rebuttal and small changes

Recently, several papers have addressed the uncertainty in BEM model output by performing uncertainty propagation and sensitivity studies, e.g.\ \cite{Echeverria2017,Matthaus2017,Murcia2018,Robertson2018,Bos2019c}. In these studies, the focus in mainly on uncertainties in the external conditions (wind parameters) and/or uncertainty in the turbine specification (geometric parameters). Apart from understanding how uncertainty in the output is related to the different uncertain inputs, such sensitivity studies are very useful to reduce the number of parameters as needed for example in design optimization \cite{Echeverria2017}. However, the uncertainty in BEM model output as caused by uncertainty in the model formulation itself, e.g.\ through the values chosen for model parameters, has been given little attention. % (exceptions being the effect of aerodynamic properties studied in \cite{Bortolotti2019,Matthaus2017})\todo{this is also what we do}.

@@

There are two ways to obtain perturbed curves: independently perturbing each of the control points (local perturbation) or using a single random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch3 - Ch7 and twist control points Tw2 - Tw7 control points, see Fig.\ \ref{perturbed_samples} (a) - (b), respectively. The twist and chord are perturbed following equation \eqref{eqn:perturb_c}, whereas the lift and drag are perturbed according to equation \eqref{eqn:perturb_cl}. \hl{Note that the multiplication of $(1+\Delta_i)$ with the unperturbed value (equations \eqref{eqn:perturb_c}, \eqref{eqn:perturb_cl}) has some physical intuition (especially for the lift and drag curves), but causes relatively large input uncertainty for the chord at the inboard sections. Similarly, we tend to obtain small perturbations for control points with values close to zero.} Examples of the resulting lift and drag curves are shown in Fig.\ \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this purpose. 
\begin{figure}[ht!]
\centering
\includegraphics[trim={0.5cm 0cm 0 0cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

@@

\subsection{Sensitivity analysis results}
In Fig.~\ref{fig:sensitivity_analysis_power} and \ref{fig:sensitivity_analysis_force}, we show the total order Sobol indices as a measure of the sensitivity of the different geometric and model parameters on the power output and the axial force, respectively, computed using LARS with $M=19$ uncertain parameters chosen from Fig.~\ref{perturbed_samples}. We found that $N=1024$ model evaluations were sufficient to achieve converged indices. In this case, we observe that the uncertainty in model parameters is at least as important as the uncertainty in geometric parameters. Furthermore, within the geometric parameters, the chord distribution shows a significantly higher sensitivity than the twist variables. Lastly, we observe that the axial force is basically only sensitive to the lift coefficient at the outward blade sections, whereas the power is sensitive to both lift and drag coefficients, also at the inward section of the blade. We note that these findings are testcase-specific and might not generalize directly to other turbines or other operational settings. Furthermore, the reported sensitivities depend on the variance that is chosen for the uncertain parameters. For example, \hl{the large sensitivity of the power with respect to uncertainty in Ch2 as shown in figure \ref{fig:sensitivity_analysis_power} is mainly due to the fact that the effective input uncertainty is larger here than at the outboard stations because of the multiplication with the local chord value (see equation \eqref{eqn:perturb_c}). Similarly, the low sensitivity with respect to twist as apparent in figures \ref{fig:sensitivity_analysis_power}-\ref{fig:sensitivity_analysis_force} is due to the fact that the effective perturbation in the twist is rather small: around 0.5 degree at the mid-span location (see equation \eqref{eqn:perturb_c}).} In future work we will relate these perturbations to values known from design standards to arrive at more realistic predictions.

\begin{figure}[h!]
\centering
\includegraphics[trim={2cm 0cm 0 0cm},clip, scale=0.4]{SA_Power_chord_twist_Cl_Cd.eps}
\caption{Sensitivity of power output towards geometric and model uncertainties.}
\label{fig:sensitivity_analysis_power}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.5]{SA_axial_force_chord_twist_Cl_Cd.eps}
\caption{Sensitivity of axial force towards geometric and model uncertainties.}
\label{fig:sensitivity_analysis_force}
\end{figure}

@@

\section{Conclusions}\label{sec:conclusions}
In this paper we have developed a framework for global sensitivity analysis using Sobol indices computed using adaptive sparse polynomial chaos expansions. The framework can be used for high-dimensional global sensitivity analysis of both geometric and model uncertainties in BEM models. As such, the proposed framework is generic and can deal with many types of (parametric) uncertainties, of which a few were investigated in this paper: twist, chord, lift coefficient and drag coefficient. By using NURBS parameterizations of chord, twist, lift and drag as a function of the blade radius, the number of uncertainties was limited but remains considerable (19 uncertain variables). Our approach (adaptive sparse polynomial chaos expansion) is able to efficiently quantify this high-dimensional problem. The results highlight amongst others the importance of the lift coefficient, especially for the axial force prediction. The identified sensitivities will be utilized in future work to develop calibrated BEM models with built-in uncertainty estimates, as part of the \textit{WindTrue} project, and used to better understand the differences between BEM models and measurement data as studied in IEA Task 29.

>>>>

Recently, several papers have addressed the uncertainty in BEM model output by performing uncertainty propagation and sensitivity studies, e.g.\ \cite{Petrone2011b,Echeverria2017,Matthaus2017,Murcia2018,Robertson2018,Bos2019c}. In these studies, the focus in mainly on uncertainties in the external conditions (wind parameters) and/or uncertainty in the turbine specification (geometric parameters). Apart from understanding how uncertainty in the output is related to the different uncertain inputs, such sensitivity studies are very useful to reduce the number of parameters as needed for example in design optimization \cite{Echeverria2017}. However, the uncertainty in BEM model output as caused by uncertainty in the model formulation itself, e.g.\ through the values chosen for model parameters, has been given little attention. % (exceptions being the effect of aerodynamic properties studied in \cite{Bortolotti2019,Matthaus2017})\todo{this is also what we do}.

@@

There are two ways to obtain perturbed curves: independently perturbing each of the control points (local perturbation) or using a single random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch3 - Ch7 and twist control points Tw2 - Tw7 control points, see Fig.\ \ref{perturbed_samples} (a) - (b), respectively. The twist and chord are perturbed following equation \eqref{eqn:perturb_c}, whereas the lift and drag are perturbed according to equation \eqref{eqn:perturb_cl}. \hl{Note that the multiplication of $(1+\Delta_i)$ with the unperturbed value (equations \eqref{eqn:perturb_c}, \eqref{eqn:perturb_cl}) has some physical intuition (especially for the lift and drag curves), but causes relatively large input uncertainty for the chord at the inboard sections (see dashed lines in figure \ref{perturbed_samples}). Similarly, we tend to obtain small perturbations for control points with values close to zero.} Examples of the resulting lift and drag curves are shown in Fig.\ \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this purpose. 
\begin{figure}[ht!]
\centering
\includegraphics[trim={0.5cm 0cm 0 0cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

@@

\subsection{Sensitivity analysis results}
In Fig.~\ref{fig:sensitivity_analysis_power} and \ref{fig:sensitivity_analysis_force}, we show the total order Sobol indices as a measure of the sensitivity of the different geometric and model parameters on the power output and the axial force, respectively, computed using LARS with $M=19$ uncertain parameters chosen from Fig.~\ref{perturbed_samples}. We found that $N=1024$ model evaluations were sufficient to achieve converged indices. In this case, we observe that the uncertainty in model parameters is at least as important as the uncertainty in geometric parameters. Furthermore, within the geometric parameters, the chord distribution shows a significantly higher sensitivity than the twist variables. Lastly, we observe that the axial force is basically only sensitive to the lift coefficient at the outward blade sections, whereas the power is sensitive to both lift and drag coefficients, also at the inward section of the blade. We note that these findings are testcase-specific and might not generalize directly to other turbines or other operational settings. Furthermore, the reported sensitivities depend on the variance that is chosen for the uncertain parameters. For example, \hl{the large sensitivity of the power with respect to uncertainty in Ch2 as shown in figure \ref{fig:sensitivity_analysis_power} is mainly due to the fact that the effective input uncertainty is larger here than at the outboard stations because of the multiplication with the local chord value (see equation \eqref{eqn:perturb_c} and figure \ref{perturbed_samples}). Similarly, the low sensitivity with respect to twist as apparent in figures \ref{fig:sensitivity_analysis_power}-\ref{fig:sensitivity_analysis_force} is due to the fact that the effective perturbation in the twist is rather small: around 0.5 degree at the mid-span location.} In future work we will relate these perturbations to values known from design standards to arrive at more realistic predictions.

\begin{figure}[h!]
\centering
\includegraphics[trim={2cm 0cm 0 0cm},clip, scale=0.4]{SA_Power_chord_twist_Cl_Cd.eps}
\caption{Sensitivity of power output towards geometric and model uncertainties.}
\label{fig:sensitivity_analysis_power}
\end{figure}
\begin{figure}[h!]
\centering
\includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.5]{SA_axial_force_chord_twist_Cl_Cd.eps}
\caption{Sensitivity of axial force towards geometric and model uncertainties.}
\label{fig:sensitivity_analysis_force}
\end{figure}

@@

\section{Conclusions}\label{sec:conclusions}
In this paper we have developed a framework for global sensitivity analysis using Sobol indices computed using adaptive sparse polynomial chaos expansions. The framework can be used for high-dimensional global sensitivity analysis of both geometric and model uncertainties in BEM models. The proposed framework is generic and can deal with many types of (parametric) uncertainties, of which a few were investigated in this paper: twist, chord, lift coefficient and drag coefficient. By using NURBS parameterizations of chord, twist, lift and drag as a function of the blade radius, the number of uncertainties was limited but remains considerable (19 uncertain variables). Our approach (adaptive sparse polynomial chaos expansion) is able to efficiently quantify this high-dimensional problem. The results highlight amongst others the importance of the lift coefficient, especially for the axial force prediction. The identified sensitivities will be utilized in future work to develop calibrated BEM models with built-in uncertainty estimates, as part of the \textit{WindTrue} project, and used to better understand the differences between BEM models and measurement data as studied in IEA Task 29.
